Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-01-30 src/tex/ms.tex
sec IIB

\section{Introduction}
\label{sec:Introduction}
% Motivation
\Ac{GW} detections by the \ac{LVK} \citep{LIGO, Virgo, KAGRA} are now routinely used to test various aspects of Einstein's theory of \ac{GR} \citep{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}.
Among those, measurements of the basic properties of \acp{GW}, like their speed and polarization, can directly probe the fundamental symmetries of the underlying theory of gravity \citep{Will:2018bme}.
For instance, unequal propagation of \ac{GW} polarization states would reveal that spacetime is birefringent, a smoking gun for parity-odd theories like Chern-Simons gravity \citep{Jackiw:2003pm, Alexander:2009tp, Sopuerta:2009iy}.
Here we constrain the magnitude of possible amplitude birefringence using \ac{BBH} signals from the latest \ac{LVK} catalog, GWTC-3.
% GW amplitude birefringence is a property of space-time that consists of the enhancement of one GW polarization over the other as the wave propagates.

@@

% GW polarization
In \ac{GR}, \acp{GW} are comprised of two independent polarization modes, usually represented in the linear basis of plus ($+$) and cross ($\times$) states.
In the Fourier domain, they can be combined into left-handed (L) and right-handed (R) circular states \cite{Isi:2022mbx},
\begin{equation}
    h_{\mathrm{L/R}} = \frac{1}{\sqrt{2}}\left(h_+ \pm i h_\times\right)\,,
\end{equation}
where $h$ is the frequency domain strain, with the plus and minus signs for L and R respectively.
These circular modes represent eigenstates of the helicity operator (helicity $\pm2$) and posses a definite parity.
Einstein's theory, which is parity even, predicts no difference in the dynamics of these two states.

@@

In particular, \emph{amplitude} birefringence would enhance one polarization mode over the other.
To first order in theories like Chern-Simons gravity, the Fourier-domain waveform observed a comoving distance $d_C$ away from the source takes the form of
\begin{equation}
    h_\mathrm{L/R}^{\mathrm{br}}(f) =
    h_\mathrm{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm\kappa\frac{d_C}{1\mathrm{Gpc}}\frac{f}{100\mathrm{Hz}}\right)\,,
    \label{eq:waveform_modification}
\end{equation}
where the emitted waveform $h_\mathrm{L/R}^{\mathrm{GR}}$ is modified by an exponential birefringent factor to yield the observed waveform $h_\mathrm{L/R}^{\mathrm{br}}$.
The overall magnitude of this effect for a given distance and frequency $f$ is set by a dimensionless opacity parameter, $\kappa$, which encodes the intrinsic strength of the birefringence.
The emitted waveform for a given source (i.e., the waveform observed very close to the source) will generally differ from the analogous waveform predicted by \ac{GR}; however, since we expect most viable modifications to \ac{GR} to be intrinsically small, it is standard to approximate the emitted waveform by the prediction from \ac{GR} (hence the notation $h^{\rm GR}$ above).

@@

Equation \eqref{eq:waveform_modification} can be derived as the first order effect in an expansion for multiple extensions of \ac{GR}.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudo-scalar field and its derivative in Chern-Simons gravity.
Since it originates from an expansion, Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, $\left|\kappa\right| \left(d_C/1\mathrm{Gpc}\right) \left(f/100\mathrm{Hz}\right) < 1$.
Otherwise, more frequency-dependent terms would enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\mi{specify what the parameter for the expansion is, add citations, and check with a theorist}

\subsection{Effect on inclination}
% Effect on inclination PE results
In GR, the amplitude ratio of the left-handed and right-handed polarizations only depends on the inclination for the dominant $(2,\pm 2)$ angular mode of nonprecessing binaries.
The relationship between the amplitude ratio and the inclination is
\begin{equation}
    \frac{h_\mathrm{L}}{h_\mathrm{R}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2\,,
\end{equation}
where $\iota$ is the inclination, the angle between our line of sight and the orbital angular momentum of the binary.
Birefringence can alter the ratio between polarizations of the waveform.
Thus, the birefringence effect could affect the PE result on $\iota$ based on \citet{Okounkova_2022}.

Based on the modification we implemented, the amplitude ratio of the left-handed and right-handed polarizations would be given by
\begin{equation}
    \frac{h_\mathrm{L_{obs}}}{h_\mathrm{R_{obs}}}=\frac{\left(1-\cos\iota\right)^2}{\left(1+\cos\iota\right)^2}\exp\left({2\kappa\frac{d_C}{1\mathrm{Gpc}}\frac{f}{100\mathrm{Hz}}}\right)\,,
    \label{eq:modified_amplitude_ratio}
\end{equation}
where $h_\mathrm{L_{obs}}$ and $h_\mathrm{R_{obs}}$ are the observed amplitude of left-handed and right-handed polarizations of the GWs.
As the amplitude ratio does not only depends on $\iota$ with this modification but also $\kappa$ and $d_C$, the PE results on $\iota$ and $d_C$ are affected.

\citet{Okounkova_2022} assume the effect of birefringence is independent of the frequency, which is a zeroth-order approximation of the birefringence model in Chern-Simons gravity.
This assumption would create a degeneracy between $\kappa$ and $\iota$, as they can affect the amplitude ratio similarly.
To reconstruct the amplitude ratio from the interferometer data, an $\iota$ representing a more face-off inspiral can pair with a positive $\kappa$, or an $\iota$ representing a more face-on inspiral with a negative $\kappa$.

In contrast, we included the frequency dependence in the equation \ref{eq:modified_amplitude_ratio}, which is a first-order approximation of the birefringence model.
The frequency dependence can break the degeneracy between $\kappa$ and $\iota$, as this would make the effect of the birefringence affect the amplitude ratio differently compared to $\iota$.
To illustrate the frequency dependence of amplitude birefringence, the observed amplitudes of two polarizations of a simulated GW signal are shown in figure \ref{fig:birefringence}.

\begin{figure}[h]
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        The observed Fourier amplitude of two polarizations of a simulated GW signal with and without birefringence.
        The amplitudes, as GR predicts, are shown in dotted lines with blue and orange colors representing the left-handed and right-handed polarizations.
        The amplitudes with birefringence are shown in solid lines with blue and orange colors representing the left-handed and right-handed polarizations.
        This plot shows the effect of amplitude birefringence with frequency dependence.
    }
    \label{fig:birefringence}
\end{figure}

>>>>

\section{Introduction}
\label{sec:Introduction}
% Motivation
\Ac{GW} detections by the \ac{LVK} \citep{LIGO, Virgo, KAGRA} are now routinely used to test various aspects of Einstein's theory of \ac{GR} \citep{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio}.
Among those, measurements of the basic properties of \acp{GW}, like their speed and polarization, can directly probe the fundamental symmetries of the underlying theory of gravity \citep{Will:2018bme}.
For instance, unequal propagation of \ac{GW} polarization states would reveal that spacetime is birefringent, a smoking gun for parity-odd theories like Chern-Simons gravity \citep{Jackiw:2003pm,Alexander:2009tp,Sopuerta:2009iy}.
Here we constrain the magnitude of possible amplitude birefringence using \ac{BBH} signals from the latest \ac{LVK} catalog, GWTC-3.
% GW amplitude birefringence is a property of space-time that consists of the enhancement of one GW polarization over the other as the wave propagates.

@@

% GW polarization
In \ac{GR}, \acp{GW} are comprised of two independent polarization modes, usually represented in the linear basis of plus ($+$) and cross ($\times$) states.
In the Fourier domain, they can be combined into left-handed (L) and right-handed (R) circular states \cite{Isi:2022mbx},
\begin{equation}
    h_{{L/R}} = \frac{1}{\sqrt{2}}\left(h_+ \pm i h_\times\right)\,,
\end{equation}
where $h$ is the frequency domain strain, with the plus and minus signs for L and R respectively.
These circular modes represent eigenstates of the helicity operator (helicity $\pm2$) and posses a definite parity.
Einstein's theory, which is parity even, predicts no difference in the dynamics of these two states.

@@

In particular, \emph{amplitude} birefringence would enhance one polarization mode over the other.
To first order in theories like Chern-Simons gravity, the Fourier-domain waveform observed a comoving distance $d_C$ away from the source takes the form of
\begin{equation}
    h_{L/R}^{\mathrm{br}}(f) =
    h_{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm\kappa\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\,\mathrm{Hz}}\right)\,,
    \label{eq:waveform_modification}
\end{equation}
where the emitted waveform $h_{L/R}^{\mathrm{GR}}$ is modified by an exponential birefringent factor to yield the observed waveform $h_{L/R}^{\mathrm{br}}$.
The overall magnitude of this effect for a given distance and frequency $f$ is set by a dimensionless opacity parameter, $\kappa$, which encodes the intrinsic strength of the birefringence.
The emitted waveform for a given source (i.e., the waveform observed very close to the source) will generally differ from the analogous waveform predicted by \ac{GR}; however, since we expect most viable modifications to \ac{GR} to be intrinsically small, it is standard to approximate the emitted waveform by the prediction from \ac{GR} (hence the notation ``$h^{\rm GR}$'' above).

@@

Equation \eqref{eq:waveform_modification} can be derived as the first order effect in an expansion for multiple extensions of \ac{GR}.
In general, $\kappa$ will be a function of the theory parameters and the cosmological history, e.g., the value of the pseudo-scalar field and its derivative in Chern-Simons gravity.
Since it originates from an expansion, Eq.~\eqref{eq:waveform_modification} is a good approximation only for small exponents, 
\begin{equation}
\left|\kappa\right| \left(d_C/1\,\mathrm{Gpc}\right) \left(f/100\, \mathrm{Hz}\right) < 1\, .
\end{equation}
Otherwise, more frequency-dependent terms would enter the exponent of Eq.~\eqref{eq:waveform_modification} in a theory-dependent way.
\mi{specify what the parameter for the expansion is, add citations, and check with a theorist}

\subsection{Relation to inclination}

% Effect on inclination PE results
Under certain conditions, the effect of birefringence can be degenerate with a change in the orientation of the source with respect to the line of sight.
Concretely, for a nonprecessing compact binary inspiral in \ac{GR}, the observed amplitude ratio of the left-handed and right-handed polarizations is only a function of the inclination $\iota$, the angle between the orbital angular momentum of the source and the line of sight.
For  the dominant $\ell = |m| = 2$ angular mode of the radiation, the relation between the amplitude ratio and the inclination is
\begin{equation}
    \frac{h_{L}^\mathrm{GR}}{h^\mathrm{GR}_{R}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2\,
\end{equation}
for all frequencies (see, e.g., Sec.~IIIC in \cite{Isi:2022mbx}).

Since birefringence impacts the observed amplitude ratio of left- and right-handed modes, it could also affect inferences about the source inclination \cite{Alexander:2009tp}.
However, the two effects are degenerate only if the frequency dependence of Eq.~\eqref{eq:waveform_modification} is neglected.
This is easy to see from Eq.~\eqref{eq:waveform_modification}, since the implied polarization ratio for the $\ell = |m| = 2$ mode of a nonprecessing source is
\begin{equation}
    \frac{h_{L}^\mathrm{br}}{h_{R}^\mathrm{br}}=\left(\frac{1-\cos\iota}{1+\cos\iota}\right)^2
    \exp\left({2\kappa\frac{d_C}{1\, \mathrm{Gpc}}\frac{f}{100\, \mathrm{Hz}}}\right)\, .
    \label{eq:modified_amplitude_ratio}
\end{equation}
For an isolated Fourier mode of definite frequency $f$, the effect of birefringence will be degenerate with a change in inclination; however, if multiple modes come into play, then no change in inclination can mask the effect of birefringence, which will generate dephasing in the time domain waveform (Fig~\ref{fig:birefringence}).

\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        The observed Fourier amplitude of two polarizations of a simulated GW signal with and without birefringence.
        The amplitudes, as GR predicts, are shown in dotted lines with blue and orange colors representing the left-handed and right-handed polarizations.
        The amplitudes with birefringence are shown in solid lines with blue and orange colors representing the left-handed and right-handed polarizations.
        This plot shows the effect of amplitude birefringence with frequency dependence.
    }
    \label{fig:birefringence}
\end{figure}

\citet{Okounkova_2022} took the effect of birefringence to be independent of the frequency, which is a zeroth-order approximation of the birefringence model in Chern-Simons gravity.
This assumption results in a full degeneracy between $\kappa$ and $\iota$:
to reconstruct the amplitude ratio from the interferometer data, an $\iota$ representing a more face-off inspiral can pair with a positive $\kappa$, or an $\iota$ representing a more face-on inspiral with a negative $\kappa$.
That fact can be used to constrain frequency-independent birefringence by searching for features in the distribution of inferred inclinations \cite{Okounkova_2022}.

By implementing Eq.~\eqref{eq:modified_amplitude_ratio}, which is a first-order approximation of the birefringence model, we generally break the degeneracy between birefringence and source orientation; this was also the case in the frequency-dependent relations studied in \cite{Yamada_2020,Wang_2021}.
Nevertheless, there exist systems for which the degeneracy cannot be perfectly broken because not enough frequencies are available in the data.
As we will see in Sec.~\ref{sec:Results}, this is the case for heavy \acp{BBH}, which are only in the sensitive band of the detectors for a few cycles.
